---
title: "Lab 10 - Landsat Imagery"
subtitle: "ECOL 620 - Applications in Landscape Ecology"
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    # code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '../';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'index.html'));
    file.copy(from = "..//index.html", to = '..///data//lab10_george_woolsey.html', overwrite = TRUE)
  })
---

# Rules

1. Using the provided R Markdown template, answer all questions and show your R code where necessary. Note, some questions are just plain text written answers.

2. Complete your assignment using the R markdown file and submit individual assignments to Canvas. Knit your script and submit an .html file on Canvas. Please use  the following naming convention: lab5_firstname_lastname.html (ex. lab7_kyle_horton.html).  Note, we will not grade labs in any other format.

# Objective

In this lab, we explore the use of Landsat 8 imagery to calculate true color images, false color images, and NDVI. We will first work with a small dataset, prepared by [Earth Lab](https://www.earthdatascience.org/), exploring the impacts of the 2016 Cold Stone fire. The Cold Stone fire was first reported on July 9, 2016 approximately two miles northeast of Nederland, Colorado. The fire was extinguished on July 14,2016, but not before burning 528 acres.

For your assignment, you will explore recent Landsat 8 data examining the pre-fire (October 2019) and post-fire (October 2020) periods surrounding the Cameron Peak fire of 2020. These data were downloaded from [Earth Explorer](https://earthexplorer.usgs.gov) at WRS row 32 and WRS path 34 (e.g., northern Colorado). You can find a KML file describing Landsat WRS rows and paths here: <https://www.usgs.gov/media/files/landsat-wrs-2-scene-boundaries-kml-file>.

---

# Setup

```{r setup, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results='hide'
  , fig.width = 10
  , fig.height = 7
)
```

```{r pkg-ld}
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(kableExtra)
library(cowplot)
library(ggtext) # color text on ggplot
# spatial analysis
library(terra)
library(sf)
library(stars)
library(starsExtra)
# set seed
set.seed(11)
```

---

# Data Preparation

Data can be downloaded from this [dropbox](https://www.dropbox.com/sh/dfqkohgr9oh9b87/AABw0t_85fQGGu9WjHyF_QkDa?dl=0)

# Question 1

Examine the files in the Cameron Peak “landsat” folder. In each folder, you’ll notice there are seven “.tif” files. Please describe the naming convention of the files. See the following for reference: <https://on.doi.gov/3d3D8wx>. (4 pts)

```{r flist, results='hide'}
# landsat folder
f_list <- list.dirs("../data/cameron_peak/landsat/", recursive = F) |> 
  purrr::map(list.files)

paste0(f_list[[1]][1],", ..., ", f_list[[1]][length(f_list[[1]])])
```

The **`r length(f_list[[1]])`** files included in the `landsat` folder include **`r paste0(f_list[[1]][1],", ..., ", f_list[[1]][length(f_list[[1]])])`**. The naming convention is xxxxxxxxxxxxxxxxxxxx

xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

# Question 2

Describe the resolution, sampling frequency, and surface reflectance bands collected by Landsat 8. (5 pts)

xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx
see RS class
xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

# Question 3

What is a World Reference System (WRS) row and WRS path? (4 pts)

xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx


# Question 4

Generate a pre- and post-fire natural color image using the Landsat tiles within the “landsat” folder. (See “Q4_figure.png”) (10 pts)

```{r ld-dta, fig.show='hide', results='hide'}
# date list
img_date_list <- 1:length(f_list) |>
  purrr::map(function(x){
    f_list[[x]][1] |> 
      toupper() |> 
      stringr::str_remove(".TIF") |> 
      stringr::word(4,sep = "_")
  }) |> 
  unlist() |> 
  lubridate::as_date()
# load data
landsat_rast_list <- 1:length(list.dirs("../data/cameron_peak/landsat/", recursive = F)) |> 
  purrr::map(function(x){
    # paths
    pth <- list.dirs("../data/cameron_peak/landsat/", recursive = F)[x]
    fls <- list.files(pth)
    rstr <- fls |> 
      purrr::map(function(f){terra::rast(paste0(pth,"/",f))}) |> 
        setNames(
          fls |> 
            toupper() |> 
            stringr::str_remove(".TIF") |> 
            stringr::word(-1,sep = "_") |> 
            tolower()
        ) |> 
        terra::rast()
    return(rstr)
  })
# plot # true color
landsat_rast_list[[1]] |> 
  terra::aggregate(fact = 6, fun = "mean", na.rm = T) |> 
  terra::plotRGB(r = 4, g = 3, b = 2, stretch = "hist", colNA = "transparent")
# plot # false color
landsat_rast_list[[2]] |> 
  terra::aggregate(fact = 6, fun = "mean", na.rm = T) |> 
  terra::plotRGB(r = 5, g = 4, b = 3, stretch = "hist", colNA = "transparent")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Plot RGB Landsat Image

```{r plt-rgb, results='hide', fig.width=12,fig.height=6}
# function to plot rgb
plt_landsat_rgb <- function(rast_list_pos) {
  # convert to stars
  landsat_img_st <- landsat_rast_list[[rast_list_pos]] |> 
    terra::subset(subset = c("b2", "b3", "b4")) |> 
    terra::aggregate(fact = 5, fun = "mean", na.rm = T) |> 
    terra::project("epsg:4326") |> 
    stars::st_as_stars()
  gc()
  # convert to rgb
  landsat_img_rgb <- stars::st_rgb(
    landsat_img_st[,,,3:1]
    , dimension = 3
    , use_alpha = FALSE
    , stretch = "histogram"
    # , probs = c(0.005, 0.995)
    # , stretch = "percent"
  )
  # ggplot
  plt_rgb <- ggplot() +
    stars::geom_stars(data = landsat_img_rgb[]) +
    scale_fill_identity(na.value = "transparent") + # !!! don't take this out or RGB plot will kill your computer
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(
      x = "Longitude"
      , y = "Latitude"
      , title = paste0(
        "Landsat Image Date: "
        , img_date_list[rast_list_pos]
      )
      , subtitle = "northern Colorado, USA (WRS row 32, WRS path 34)"
    ) +
    theme_light() +
    theme(
      legend.position = "none"
      , axis.title = element_text(size = 8)
      , axis.text = element_text(size=7)
      , panel.grid = element_blank()
      , plot.title = element_text(size = 10, hjust = 0.5, face="bold")
      , plot.subtitle = element_text(size = 8, hjust = 0.5)
    )
  return(plt_rgb)
}
# map function
plt_list <- 1:length(landsat_rast_list) |> 
  purrr::map(plt_landsat_rgb)
# combine
cowplot::plot_grid(
  plotlist = plt_list
  , nrow = 1
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```


# Question 5

Generate a three-panel plot showing a false color image (bands 7-5-4) of the pre-fire, post-fire, and severity raster (“CameronPeak_SBS_final.tif”) cropped to the Cameron Peak area. Hint, crop to the following extent: extent(400000, 500000, 4445400, 4545000). (See “Q5_figure.png”) (10 pts)

```{r}

```


# Question 6

Generate an NDVI surface for the post-fire (October 6th, 2020) period cropped to the Fort Collins area. Hint, crop to the following extent: extent(482511.8, 509752.3, 4482326, 4498192). ). (See “Q6_figure.png”) (10 pts)

```{r}

```


# Question 7

From the previous plot, which areas (think cover types) show the lowest NDVI values and which show the highest? (5 pts)

```{r}

```

<span style="color: teal;">
asdf
</span>