---
title: "Lab 10 - Landsat Imagery"
subtitle: "ECOL 620 - Applications in Landscape Ecology"
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    # code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '../';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'index.html'));
    file.copy(from = "..//index.html", to = '..///data//lab10_george_woolsey.html', overwrite = TRUE)
  })
---

# Rules

1. Using the provided R Markdown template, answer all questions and show your R code where necessary. Note, some questions are just plain text written answers.

2. Complete your assignment using the R markdown file and submit individual assignments to Canvas. Knit your script and submit an .html file on Canvas. Please use  the following naming convention: lab5_firstname_lastname.html (ex. lab7_kyle_horton.html).  Note, we will not grade labs in any other format.

# Objective

In this lab, we explore the use of Landsat 8 imagery to calculate true color images, false color images, and NDVI. We will first work with a small dataset, prepared by [Earth Lab](https://www.earthdatascience.org/), exploring the impacts of the 2016 Cold Stone fire. The Cold Stone fire was first reported on July 9, 2016 approximately two miles northeast of Nederland, Colorado. The fire was extinguished on July 14,2016, but not before burning 528 acres.

For your assignment, you will explore recent Landsat 8 data examining the pre-fire (October 2019) and post-fire (October 2020) periods surrounding the Cameron Peak fire of 2020. These data were downloaded from [Earth Explorer](https://earthexplorer.usgs.gov) at WRS row 32 and WRS path 34 (e.g., northern Colorado). You can find a KML file describing Landsat WRS rows and paths here: <https://www.usgs.gov/media/files/landsat-wrs-2-scene-boundaries-kml-file>.

---

# Setup

```{r setup, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results='hide'
  , fig.width = 10
  , fig.height = 7
)
```

```{r pkg-ld}
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(kableExtra)
library(cowplot)
library(ggtext) # color text on ggplot
# spatial analysis
library(terra)
library(sf)
library(stars)
library(starsExtra)
# set seed
set.seed(11)
```

---

# Data Preparation

Data can be downloaded from this [dropbox](https://www.dropbox.com/sh/dfqkohgr9oh9b87/AABw0t_85fQGGu9WjHyF_QkDa?dl=0)

# Question 1

Examine the files in the Cameron Peak “landsat” folder. In each folder, you’ll notice there are seven “.tif” files. Please describe the naming convention of the files. See the following for reference: <https://on.doi.gov/3d3D8wx>. (4 pts)

```{r flist, results='hide'}
# landsat folder
f_list <- list.dirs("../data/cameron_peak/landsat/", recursive = F) |> 
  purrr::map(list.files)
```

<span style="color: teal;">
The **`r length(f_list[[1]])`** files included in the `landsat` folder include **`r paste0(f_list[[1]][1],", ..., ", f_list[[1]][length(f_list[[1]])])`**. The Landsat product identifier (e.g. `r f_list[[1]][1]`) is composed of the Landsat Collection processing levels, processing date, collection number, and collection tier category. The product identifier follows the convention:
</span>
 
*LXSS_LLLL_PPPRRR_YYYYMMDD_yyyymmdd_CC_TX*

Where:

* L = Landsat
* X = Sensor (“C”=OLI/TIRS combined, “O”=OLI-only, “T”=TIRS-only, “E”=ETM+, “T”=“TM, “M”=MSS). Where, Operational Land Image (OLI) and Thermal Infrared Sensor (TIRS).
* SS = Satellite (”07”=Landsat 7, “08”=Landsat 8)
* LLL = Processing correction level (L1TP/L1GT/L1GS)
* PPP = WRS path
* RRR = WRS row
* YYYYMMDD = Acquisition year, month, day
* yyyymmdd - Processing year, month, day
* CC = Collection number (01, 02, …)
* TX = Collection category (“RT”=Real-Time, “T1”=Tier 1, “T2”=Tier 2)

<span style="color: teal;">
Therefore, the image that was used above "LC08_L2SP_034032_20191004_20200825_02_T1_SR_B1" means: Landsat 8; OLI/TIRS combined; processing correction level L2SP; WRS path 034; WRS row 032; acquired October 04, 2019; processed August 25, 2020; Collection 2; Tier 1; Surface Reflectance values; Band 1.
</span>


# Question 2

Describe the resolution, sampling frequency, and surface reflectance bands collected by Landsat 8. (5 pts)

<span style="color: teal;">
Landsat 8 has 11 spectral bands each covering different wavelengths and at varying spatial resolution as listed in the table below. The spatial resolution ranges from a low of 15m for the Panchromatic band (band 8) to a high of 100m for the Thermal Infrared bands (band 10 and band 11). The remaining bands have a spatial resolution of 30m. The temporal resolution of Landsat 8 imagery is 16 days.
</span>

```{r, include=FALSE, eval=TRUE}
data.frame(
  band = c(1,2,3,4,5,6,7,8,9,10,11)
  , desc = c("Coastal aerosol","Blue","Green","Red","Near Infrared (NIR)"
             ,"SWIR 1","SWIR 2","Panchromatic","Cirrus"
             ,"Thermal Infrared (TIRS) 1","Thermal Infrared (TIRS) 2")
  , wavelength = c("0.433–0.453","0.450–0.515","0.525–0.600","0.630–0.680","0.845–0.885","1.560–1.660","2.100–2.300","0.500–0.680","1.360–1.390","10.6-11.2","11.5-12.5")
  , resolution = c("30 m","30 m","30 m","30 m","30 m","30 m","30 m","15 m","30 m","100 m","100 m")
) |> 
kableExtra::kable(
  format = "html" 
  , caption = "Landsat 8 Band Description"
  , col.names = c(
    "Band Number","Description","Wavelength (µm)","Resolution (m)"
  )
) %>% 
kable_styling(font_size = 11)
```


# Question 3

What is a World Reference System (WRS) row and WRS path? (4 pts)

<span style="color: teal;">
The Worldwide Reference System (WRS) is a positioning system for Landsat imagery. The WRS row is analogous to the latitudinal center line of an image. The WRS path numbers are ordered sequentially from east to west based on the track of the satellite as it orbits the Earth. The WRS path number and WRS row number are combined in the formate "path-row" to uniquely identify the scene center of a Landsat image (e.g. "034-032" relates to WRS path 034; WRS row 032).
</span>

# Question 4

Generate a pre- and post-fire natural color image using the Landsat tiles within the “landsat” folder. (See “Q4_figure.png”) (10 pts)

```{r ld-dta, fig.show='hide', results='hide'}
# date list
img_date_list <- 1:length(f_list) |>
  purrr::map(function(x){
    f_list[[x]][1] |> 
      toupper() |> 
      stringr::str_remove(".TIF") |> 
      stringr::word(4,sep = "_")
  }) |> 
  unlist() |> 
  lubridate::as_date()
# load data
landsat_rast_list <- 1:length(list.dirs("../data/cameron_peak/landsat/", recursive = F)) |> 
  purrr::map(function(x){
    # paths
    pth <- list.dirs("../data/cameron_peak/landsat/", recursive = F)[x]
    fls <- list.files(pth)
    rstr <- fls |> 
      purrr::map(function(f){terra::rast(paste0(pth,"/",f))}) |> 
        setNames(
          fls |> 
            toupper() |> 
            stringr::str_remove(".TIF") |> 
            stringr::word(-1,sep = "_") |> 
            tolower()
        ) |> 
        terra::rast()
    return(rstr)
  })
# plot # true color
landsat_rast_list[[1]] |> 
  terra::aggregate(fact = 6, fun = "mean", na.rm = T) |> 
  terra::plotRGB(r = 4, g = 3, b = 2, stretch = "hist", colNA = "transparent")
# plot # false color
landsat_rast_list[[2]] |> 
  terra::aggregate(fact = 6, fun = "mean", na.rm = T) |> 
  terra::plotRGB(r = 5, g = 4, b = 3, stretch = "hist", colNA = "transparent")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Plot RGB Landsat Image

```{r plt-rgb, results='hide', fig.width=12,fig.height=6}
# function to plot rgb
plt_landsat_three_bands <- function(
    rast_list_pos
    , three_band_list = c("b4", "b3", "b2")
    , my_fact = 7
    , my_ext = terra::ext(landsat_rast_list[[1]])
  ) {
  # convert to stars
  landsat_img_st <- landsat_rast_list[[rast_list_pos]] |> 
    terra::subset(subset = three_band_list) |> 
    terra::crop(my_ext) |> 
    terra::aggregate(fact = my_fact, fun = "mean", na.rm = T) |> 
    terra::project("epsg:4326") |> 
    stars::st_as_stars()
  gc()
  # convert to rgb
  landsat_img_rgb <- stars::st_rgb(
    landsat_img_st[,,,1:3]
    , dimension = 3
    , use_alpha = FALSE
    , stretch = "histogram"
    # , probs = c(0.005, 0.995)
    # , stretch = "percent"
  )
  # ggplot
  plt_rgb <- ggplot() +
    stars::geom_stars(data = landsat_img_rgb[]) +
    scale_fill_identity(na.value = "transparent") + # !!! don't take this out or RGB plot will kill your computer
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(
      x = "Longitude"
      , y = "Latitude"
      , title = paste0(
        "Landsat Image Date: "
        , img_date_list[rast_list_pos]
      )
      , subtitle = "northern Colorado, USA (WRS row 32, WRS path 34)"
    ) +
    theme_light() +
    theme(
      legend.position = "none"
      , axis.title = element_text(size = 8)
      , axis.text = element_text(size=7)
      , panel.grid = element_blank()
      , plot.title = element_text(size = 10, hjust = 0.5, face="bold")
      , plot.subtitle = element_text(size = 8, hjust = 0.5)
    )
  return(plt_rgb)
}
# map function
plt_list_rgb <- 1:length(landsat_rast_list) |> 
  purrr::map(plt_landsat_three_bands, my_fact = 5)
# combine
cowplot::plot_grid(
  plotlist = plt_list_rgb
  , nrow = 1
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```


# Question 5

Generate a three-panel plot showing a false color image (bands 7-6-4) of the pre-fire, post-fire, and severity raster (“CameronPeak_SBS_final.tif”) cropped to the Cameron Peak area. Hint, crop to the following extent: extent(400000, 500000, 4445400, 4545000). (See “Q5_figure.png”) (10 pts)

## False color plots

```{r plt-false, results='hide', fig.show='hide'}
# map function
plt_list_false <- 1:length(landsat_rast_list) |> 
  purrr::map(
    .f = plt_landsat_three_bands
    , three_band_list = c("b7","b6","b4")
    , my_fact = 3
    , my_ext = c(400000, 500000, 4445400, 4545000)
  )
# combine
cowplot::plot_grid(
  plotlist = plt_list_false
  , nrow = 1
)
```

## Severity plot

```{r plt-severity, results='hide'}
# load severity map
cp_severity <- terra::rast("../data/cameron_peak/fire_severity/CameronPeak_SBS_final.tif") |> as.factor()
# temp rgb
landsat_img_temp <- landsat_rast_list[[2]] |> 
  terra::subset(subset = c("b4", "b3", "b2")) |> 
  terra::crop(c(400000, 500000, 4445400, 4545000)) |>
  terra::aggregate(fact = 3, fun = "mean", na.rm = T) |> 
  terra::project("epsg:4326") |> 
  stars::st_as_stars()
gc()
# plot
plt_severity <- 
  ggplot() +
    stars::geom_stars(
      data = stars::st_rgb(
            landsat_img_temp[,,,1:3]
            , dimension = 3
            , use_alpha = FALSE
            , stretch = "histogram"
          )
    ) +
    scale_fill_identity(na.value = "transparent") + # !!! don't take this out or RGB plot will kill your computer
    ggnewscale::new_scale_fill() +
    geom_raster(
      data = cp_severity |> 
          terra::crop(c(400000, 500000, 4445400, 4545000)) |>
          terra::project("epsg:4326") |> 
          as.data.frame(xy = T) |>
          dplyr::rename(severity = 3) |> 
          dplyr::filter(severity > 0) |> 
          dplyr::mutate(
            severity_fact = factor(
                severity
                , levels = 1:4
                , labels = c("unburned", "low", "moderate", "high")
                , ordered = T
              )
          )
          
      , mapping = aes(x=x,y=y,fill=severity_fact)
    ) +
    scale_fill_brewer(
      type = "div"
      , palette = "RdYlBu"
      , na.value = "transparent"
      , direction = -1
    ) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(
      x = "Longitude"
      , y = "Latitude"
      , fill = "severity"
      , title = "Wildfire Burn Severity"
      , subtitle = " "
    ) +
    theme_light() +
    theme(
      legend.position = c(0.5,1)
      , legend.direction = "horizontal"
      , legend.text = element_text(size = 8)
      , legend.title = element_blank()
      , axis.title = element_text(size = 8)
      , axis.text = element_text(size=7)
      , panel.grid = element_blank()
      , plot.title = element_text(size = 10, hjust = 0.5, face="bold")
      , plot.subtitle = element_text(size = 8, hjust = 0.5)
    )
```

## Combine plots

```{r plt-cmbn-fls-sev, fig.height=18, results='hide'}
plt_list_false[[length(plt_list_false)+1]] = plt_severity
cowplot::plot_grid(
  plotlist =  plt_list_false
  , ncol = 1
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```


# Question 6

Generate an NDVI surface for the post-fire (October 6th, 2020) period cropped to the Fort Collins area. Hint, crop to the following extent: extent(482511.8, 509752.3, 4482326, 4498192). ). (See “Q6_figure.png”) (10 pts)


```{r plt-ndvi}
# extent
crp_ext_temp <- c(482511.8, 509752.3, 4482326, 4498192)
# pull red and infrared bands
red_temp <- landsat_rast_list[[2]] |> 
  terra::subset(subset = c("b4")) |> 
  terra::crop(crp_ext_temp)
infrared_temp <- landsat_rast_list[[2]] |> 
  terra::subset(subset = c("b5")) |> 
  terra::crop(crp_ext_temp)
# calculate ndvi
ndvi_rast <- ((infrared_temp - red_temp) / (infrared_temp + red_temp)) |> 
  terra::aggregate(fact = 2, fun = "mean", na.rm = T) |> 
  terra::project("epsg:4326")
# plot
ndvi_rast |> 
  as.data.frame(xy = T) |> 
  dplyr::rename(ndvi=3) |> 
  ggplot() + 
    geom_raster(mapping=aes(x=x,y=y,fill=ndvi)) +
    scale_fill_distiller(
      type = "seq"
      , palette = "YlGn"
      , direction = 1
      # , limits = c(-1,1)
      , na.value = "transparent"
    ) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(
      x = "Longitude"
      , y = "Latitude"
      , fill = "NDVI"
      , title = paste0(
          "Landsat Image Date: "
          , img_date_list[2]
        )
      , subtitle = "Fort Collins, CO Area NDVI"
    ) +
    theme_light() +
    theme(
      legend.position = c(0.05,0.15)
      , legend.text = element_text(size = 8)
      , legend.title = element_text(size = 8)
      , axis.title = element_text(size = 8)
      , axis.text = element_text(size=7)
      , panel.grid = element_blank()
      , plot.title = element_text(size = 10, hjust = 0.5, face="bold")
      , plot.subtitle = element_text(size = 8, hjust = 0.5)
    )

```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

# Question 7

From the previous plot, which areas (think cover types) show the lowest NDVI values and which show the highest? (5 pts)

<span style="color: teal;">
The areas that have the highest NDVI values (in this image values near +0.5) correspond to areas with healthy, green vegetation. The areas with low NDVI values (in this image values near -0.1) correspond to areas with little or no healthy vegetation. These low-NDVI areas include bodies of water such as Horsetooth Reservoir (40.56°, -105.16°).
</span>
