---
title: "Lab 10 - Landsat Imagery"
author: "ECO620  - Applications in Landscape Ecology"
date: "Spring 2023"
output: 
  html_document:
    toc: true
    theme: united 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.align='center', 
                      message = FALSE, 
                      warning = FALSE,
                      fig.pos = "H")
```

## Environment set up

Download lab files from [this folder](https://www.dropbox.com/sh/dfqkohgr9oh9b87/AABw0t_85fQGGu9WjHyF_QkDa?dl=0).

For more information about regular expressions, look here: <https://en.wikipedia.org/wiki/Regular_expression>

It is a tool frequently used to select files with names that match a certain pattern of interest (though it does an infinite number of other things).

```{r}

# Load spatial packages
library(raster)
library(rgdal)
library(rgeos)
#library(RStoolbox)
library(ggplot2)
library(scales)
library(cowplot)

# Set up path to directory where lab files are stored:
dir_files = "./data_for_lab10/coldstone"

# Set up path to directories with post and pre fire files:
pre_fire_dir = paste(dir_files, "LC80340322016189-SC20170128091153", sep="/")
post_fire_dir = paste(dir_files, "LC80340322016205-SC20170127160728", sep="/")

# List all the band files in pre and post fire directories using regular 
# expressions to select band files:
pre_band_files <- list.files(pre_fire_dir,
                                pattern = glob2rx("*band*.tif$"),
                                full.names = TRUE) 
post_band_files <- list.files(post_fire_dir,
                                pattern = glob2rx("*band*.tif$"),
                                full.names = TRUE)


```

# Inspecting a band file (example)

```{r, fig.width=5}

# Let`s load the second file in the pre-fire list and see what is in it:
pre_band2 <- raster(pre_band_files[2])

plot(pre_band2, main = "Landsat 8 cropped band 2\nCold Springs fire scar",
     col = gray(0:100 / 100))

```

# What are all the bands? How to load them?

| Band                         | Wavelength range (nanometers) | Spatial Resolution (m) | Spectral Width (nm) |
|------------------------------|-------------------------------|------------------------|---------------------|
| Band 1 - Coastal aerosol     | 430 - 450                     | 30                     | 2.0                 |
| Band 2 - Blue                | 450 - 510                     | 30                     | 6.0                 |
| Band 3 - Green               | 530 - 590                     | 30                     | 6.0                 |
| Band 4 - Red                 | 640 - 670                     | 30                     | 0.03                |
| Band 5 - Near Infrared (NIR) | 850 - 880                     | 30                     | 3.0                 |
| Band 6 - SWIR 1              | 1570 - 1650                   | 30                     | 8.0                 |
| Band 7 - SWIR 2              | 2110 - 2290                   | 30                     | 18                  |
| Band 8 - Panchromatic        | 500 - 680                     | 15                     | 18                  |
| Band 9 - Cirrus              | 1360 - 1380                   | 30                     | 2.0                 |

### Creating a stack of all band files

```{r}

# Stack the data (check the class of the object created and notice what the stack function did):
pre_bands <- stack(pre_band_files)
post_bands <- stack(post_band_files)

# view stack attributes
pre_bands

```


```{r}

# Get list of each layer name
names(pre_bands)

# Remove the filename from each band name for pretty plotting:
names(pre_bands) <- gsub(pattern = "LC80340322016189LGN00_sr_", replacement = "", names(pre_bands))
names(post_bands) <- gsub(pattern = "LC80340322016205LGN00_sr_", replacement = "", names(post_bands))

# Check the result:
names(pre_bands)

```

### Plot ugly panel of all bands

```{r, fig.width = 6, fig.height = 8}

plot(pre_bands,
     col = gray(20:100 / 100),
     nr = 4, nc = 2)

```

# Creating RGB composite plot

```{r}

# Subset red, green, and blue color bands:
pre_rgb = raster::subset(pre_bands, c(2, 3, 4)) 

# Rescale the values to have min = 0 and max = 255 (256 color scale):
pre_rgb_scaled <- stretch(x=pre_rgb, minv=0, maxv=255)

# Make it ggplot friendly:
pre_rgb_scaled_df = as.data.frame(pre_rgb_scaled, xy = T)

pre_rgb_plot_coldstone = ggplot()+
  geom_raster(data=pre_rgb_scaled_df, aes(x=x, y=y, 
                                          fill=rgb(blue = band2_crop, 
                                                   green = band3_crop, 
                                                   red = band4_crop, 
                                                   maxColorValue = 255)))+ 
  coord_equal(xlim = range(pre_rgb_scaled_df$x), ylim = range(pre_rgb_scaled_df$y))+ 
  theme_bw(base_size = 14) + scale_fill_identity()+
  scale_x_continuous(breaks=range(pre_rgb_scaled_df$x)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  scale_y_continuous(breaks=range(pre_rgb_scaled_df$y)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  labs(title = "Cold Stone - July 7th, 2016",
       subtitle = "Pre-fire",
       x = "Easting (meters)", y = "Northing (meters)")+
  theme(axis.line=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        plot.background=element_blank())

# Subset red, green, and blue color bands:
post_rgb = raster::subset(post_bands, c(2, 3, 4)) 

# Rescale the values to have min = 0 and max = 255 (256 color scale):
post_rgb_scaled <- stretch(x=post_rgb, minv=0, maxv=255)

# Make it ggplot friendly:
post_rgb_scaled_df = as.data.frame(post_rgb_scaled, xy = T)

post_rgb_plot_coldstone = ggplot()+
  geom_raster(data=post_rgb_scaled_df, aes(x=x, y=y, 
                                          fill=rgb(blue = band2_crop, 
                                                   green = band3_crop, 
                                                   red = band4_crop, 
                                                   maxColorValue = 255)))+ 
  coord_equal(xlim = range(post_rgb_scaled_df$x), ylim = range(post_rgb_scaled_df$y))+ 
  theme_bw(base_size = 14) + scale_fill_identity()+
  scale_x_continuous(breaks=range(post_rgb_scaled_df$x)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  scale_y_continuous(breaks=range(post_rgb_scaled_df$y)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  labs(title = "Cold Stone - July 7th, 2016",
       subtitle = "Post-fire",
       x = "Easting (meters)", y = "Northing (meters)")+
  theme(axis.line=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        plot.background=element_blank())

coldstone_grid = plot_grid(pre_rgb_plot_coldstone, post_rgb_plot_coldstone, 
                           nrow = 1, ncol = 2)

coldstone_grid

```

# Creating a false color composite

```{r}

# Subset red, green, and blue color bands:
pre_false = raster::subset(pre_bands, c(5, 6, 7)) 

# Rescale the values to have min = 0 and max = 255 (256 color scale):
pre_false_scaled <- stretch(x=pre_false, minv=0, maxv=255)

# Make it ggplot friendly:
pre_false_scaled_df = as.data.frame(pre_false_scaled, xy = T)

pre_false_plot_coldstone = ggplot()+
  geom_raster(data=pre_false_scaled_df, aes(x=x, y=y, 
                                          fill=rgb(blue = band5_crop, 
                                                   green = band6_crop, 
                                                   red = band7_crop, 
                                                   maxColorValue = 255)))+ 
  coord_equal(xlim = range(pre_false_scaled_df$x), ylim = range(pre_false_scaled_df$y))+ 
  theme_bw(base_size = 14) + scale_fill_identity()+
  scale_x_continuous(breaks=range(pre_false_scaled_df$x)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  scale_y_continuous(breaks=range(pre_false_scaled_df$y)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  labs(title = "Cold Stone - July 7th, 2016",
       subtitle = "Pre-fire (bands: 567)",
       x = "Easting (meters)", y = "Northing (meters)")+
  theme(axis.line=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        plot.background=element_blank())

# Subset red, green, and blue color bands:
post_false = raster::subset(post_bands, c(5, 6, 7)) 

# Rescale the values to have min = 0 and max = 255 (256 color scale):
post_false_scaled <- stretch(x=post_false, minv=0, maxv=255)

# Make it ggplot friendly:
post_false_scaled_df = as.data.frame(post_false_scaled, xy = T)

post_false_plot_coldstone = ggplot()+
  geom_raster(data=post_false_scaled_df, aes(x=x, y=y, 
                                          fill=rgb(blue = band5_crop, 
                                                   green = band6_crop, 
                                                   red = band7_crop, 
                                                   maxColorValue = 255)))+ 
  coord_equal(xlim = range(post_false_scaled_df$x), ylim = range(post_false_scaled_df$y))+ 
  theme_bw(base_size = 14) + scale_fill_identity()+
  scale_x_continuous(breaks=range(post_false_scaled_df$x)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  scale_y_continuous(breaks=range(post_false_scaled_df$y)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  labs(title = "Cold Stone - July 7th, 2016",
      subtitle = "Post-fire (bands: 567)",
       x = "Easting (meters)", y = "Northing (meters)")+
  theme(axis.line=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        plot.background=element_blank())

coldstone_grid = plot_grid(pre_false_plot_coldstone, post_false_plot_coldstone, 
                           nrow = 1, ncol = 2)

# Witness the magic of bands:
coldstone_grid

```

# Calculating NDVI

```{r, out.width="70%"}

red = post_bands[[4]]
infrared = post_bands[[5]]

# Perform raster algebra to calculate a raster of NDVI values
ndvi = (infrared - red) / (infrared + red)

# Plot the NDVI as a false-color image
ndvi_df = as.data.frame(ndvi, xy=T)

ggplot()+
  geom_raster(data=ndvi_df, aes(x=x, y=y, fill=layer))+
  theme(plot.title = element_text(hjust = 0.5))+
  theme_bw(base_size = 14)+ coord_equal()+
  scale_fill_gradient2()+
  scale_x_continuous(breaks=range(ndvi_df$x)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  scale_y_continuous(breaks=range(ndvi_df$y)*c(1.001, 0.999), 
                     labels=label_number(accuracy = 1), 
                     expand = c(0,0))+
  labs(title = "Cold Stone - July 7th, 2016",
       subtitle = "Post-fire NDVI",
       x = "Easting (meters)", y = "Northing (meters)",
       fill = "NDVI")+
  theme(axis.line=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        plot.background=element_blank())
```
